---
title: "Class 15: Investigating Pertussis Resurgence"
author: "Saba Heydari Seradj"
format: pdf
editor: visual
---

# Background 
Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis (Figure 1). People of all ages can be infected leading to violent coughing fits followed by a characteristic high-pitched “whoop” like intake of breath. Children have the highest risk for severe complications and death. Recent estimates from the WHO indicate that ~16 million cases and 200,000 infant deaths are due to pertussis annually (Black et al. 2010).

# Investigating pertussis cases by year

## Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

Problem... data is in pdf format. We will use datapasta package to scrape data into R as a dataframe. 

```{r}
#install.packages('datapasta')
```

```{r}
cdc <- data.frame(
                          year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L),
  cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044)
)
```

```{r}
head(cdc)
```

```{r}
library(ggplot2)
library(scales)

# Building the plot
baseplot <- ggplot(cdc) +
aes(x = year,
y = cases) +
geom_point() +
geom_line() +
labs(x = "Year",
y = "Number of cases") +
scale_y_continuous(labels = comma) # No longer scientific notation
baseplot
```

# A tale of two vaccines (wP & aP)

## Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine. What do you notice?

```{r}
# Landmark plot
lm_plot <- baseplot +
geom_vline(xintercept = 1946, # wP vaccine with everything
col = 'blue') +
geom_vline(xintercept = 1996, # aP vaccine with "essential components"
col = 'red') +
geom_vline(xintercept = 2003, # Start of the big increase
col = 'darkgreen') +
geom_vline(xintercept = c(2020,2021), # Covid-19 lockdowns
col = 'grey')
lm_plot
```

## Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

aP was introduced in 1996 and for ~10 years there was a steady state. After that, around 2004, there was a big increase in cases which could be due to various reasons, such as bacterial evolution, short-lasting effectiveness of aP, and antivax movements. 

# Exploring CMI-PB data
Problem...data is in JSON format. We will use jsonlite package to process JSON data. 

```{r}
#install.packages('jsonlite')
```

```{r}
library(jsonlite)

# Read subject table
subject <- read_json('https://www.cmi-pb.org/api/v5/subject',
simplifyVector = TRUE)

head(subject)
```

## Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 85 wP and 87 aP vaccinated subjects. 

## Q5. How many Male and Female subjects/patients are in the dataset?
```{r}
table(subject$biological_sex)
```

112 females, 60 males. 

## Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

## Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
specimens <- read_json('https://www.cmi-pb.org/api/v5/specimen',
simplifyVector = TRUE)
head(specimens)
```

```{r}
library(dplyr)
meta <- inner_join(specimens, subject)
```

```{r}
head(meta)
```

## Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- read_json('http://cmi-pb.org/api/v5/plasma_ab_titer',
simplifyVector = TRUE)
head(abdata)
```

```{r}
ab <- inner_join(abdata, meta)
```

```{r}
head(ab)
```

```{r}
nrow(ab)
```

## Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(ab$isotype)
```

Number of antigens: 

```{r}
table(ab$antigen)
```
Focusing on IgG...

```{r}
igg <- filter(ab, isotype == 'IgG')
head(igg)
```

```{r}
ggplot(igg) +
aes(MFI_normalised, antigen) +
geom_boxplot()
```

```{r}
ggplot(igg) +
aes(MFI_normalised, antigen, col = infancy_vac) +
geom_boxplot()
```

```{r}
ggplot(igg) +
aes(MFI_normalised, antigen, col = infancy_vac) +
geom_boxplot() +
facet_wrap(~visit) # Faceting by visit
```

```{r}
ggplot(igg) +
aes(MFI_normalised, antigen, col = infancy_vac) +
geom_boxplot() +
facet_wrap(~infancy_vac) # Faceting by vaccine
```

```{r}
table(igg$visit)
```

We can see a trend. THere are a lot of visitations in the beginning but decreases towards later ones. We'll focus on visits 1-7. 

```{r}
igg_7 <- filter(igg, visit %in% 1:7)
table(igg_7$visit)
```

```{r}
ggplot(igg_7) +
aes(MFI_normalised, antigen, col = infancy_vac) +
geom_boxplot() +
facet_wrap(~visit, ncol = 2) # Faceting by visit
```

```{r}
ggplot(igg_7) +
aes(x = visit,
y = MFI_normalised,
group=subject_id) +
geom_point() +
geom_line()
```

```{r}
abdata.21 <- ab %>% filter(dataset == "2021_dataset")
abdata.21 %>%
filter(isotype == "IgG", antigen == "PT") %>%
ggplot() +
aes(x=planned_day_relative_to_boost,
y=MFI_normalised,
col=infancy_vac,
group=subject_id) +
geom_point() +
geom_line()
```
