---
title: "Class 12"
author: "Saba Heydari Seradj"
format: pdf
editor: visual
---

#  Section 1. Proportion of G/G in a population 

Downloaded a CSV file from Ensemble. 

Here we read the CSV file: 
```{r}
mxl <- read.csv('genotypes.csv')
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

14.06% of the population are G|G. 

# Section 4. Population analysis

```{r}
expr <- read.table('rs8067378_ENSG00000172057.6.txt')
head(expr)
```

How many samples do we have? 
```{r}
nrow(expr)
```

462 samples. 

What is the sample size for each genotype? 
```{r}
table(expr$geno)
```

A/A has 108 samples, A/G has 233, and G/G has 121.

```{r}
summary(expr)
```
What is the median expression levels for each genotype? 

```{r}
median_expression <- tapply(expr$exp, expr$geno, median, na.rm = TRUE)
median_expression
```

Let's make a boxplot!

```{r}
library(ggplot2)

g <- ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=T)
g
```
## Q14
We can infer that the G/G SNP reduces expression of ORMDL3 as it has the lowest expression compared to other genotypes. 


