---
title: "class10"
author: "Saba Heydari Seradj (A17002175)"
format: pdf
editor: visual
---

# Importing Candy Data

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

## Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

85 different candy types. 

## Q2. How many fruity candy types are in the dataset?

```{r}
table(candy$fruity)
```

38 fruity candy types. 

## Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
rownames(candy)
```

```{r}
candy['Twix', 'winpercent']
```

My favorite candy is Twix with a win percent of 81%. 

## Find fruity candy with a winpercent above 50%

```{r}
library(dplyr)
```
```{r}
fruit_win <- candy |>
filter(winpercent > 50) |>
filter(fruity == 1)
head(fruit_win)
```

## Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy['Kit Kat', 'winpercent']
```

Win percent of kit kat is ~77%. 

## Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy['Tootsie Roll Snack Bars', 'winpercent']
```

Win percent of Tootsie Roll is ~50%. 

## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
#install.packages("skimr")
```

```{r}
library("skimr")
skim(candy)
```

Win percent is on a different scale from other variables. Need scaling. 

## Q7. What do you think a zero and one represent for the candy$chocolate column?

These are logicals. 0 means it is not a chocolate, 1 means it is a type of chocolate. 

## Q8. Plot a histogram of winpercent values


```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent) +
geom_histogram(binwidth = 6) +
theme_minimal()
```

## Q9. Is the distribution of winpercent values symmetrical?
No, it seems skewed. 

## Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

Median is ~48 which is below 50%. 

## Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate_candy <- candy |>
filter(chocolate==1)
head(chocolate_candy)
```
```{r}
fruity_candy <- candy |>
filter(fruity==1)
head(fruity_candy)
```

```{r}
summary(chocolate_candy$winpercent)
```

```{r}
summary(fruity_candy$winpercent)
```

It seems that chocolate candy has higher win percent on average (~61%) compared to fruity candy (~44%).

## Q12. Is this difference statistically significant?
We can perform a Welch's t test to test this. 
```{r}
t.test(chocolate_candy$winpercent, fruity_candy$winpercent)
```
The p-value is very small and below 0.05, so it seems the difference is statistically significant. 

# Overall Candy Rankings

## Q13. What are the five least liked candy types in this set?

```{r}
least_candy <- candy |>
arrange((winpercent))
head(least_candy, 5)
```
## Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
highest_candy <- candy |>
arrange(desc(winpercent))
head(highest_candy, 5)
```

## Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent, rownames(candy)) +
geom_col() + theme_minimal()
```

## Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
aes(winpercent,
reorder(rownames(candy), winpercent)) +
geom_col() + theme_minimal()
```
```{r}
# make color vector placeholder
my_cols <- rep('black', nrow(candy))

my_cols[as.logical(candy$chocolate)] <- 'brown'
my_cols[as.logical(candy$bar)] <- 'red'
my_cols[as.logical(candy$fruity)] <- 'orange'
```

```{r}
ggplot(candy) +
aes(winpercent,
reorder(rownames(candy), winpercent)) +
geom_col(fill=my_cols)
```

## Q17. What is the worst ranked chocolate candy?
sixlits

## Q18. What is the best ranked fruity candy?
starbursts

# Taking a look at pricepercent

```{r}
my_cols[as.logical(candy$fruity)] <- 'blue'

ggplot(candy) +
aes(x = winpercent,
y = pricepercent) +
geom_point(col = my_cols)
```

Let's add some labels: 
```{r}
#install.packages('ggrepel')
```

```{r}
library(ggrepel)

ggplot(candy) +
aes(x = winpercent,
y = pricepercent,
label=rownames(candy)) +
geom_point(col = my_cols) +
geom_text_repel(col = my_cols, max.overlaps = 8)
```

## Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's mini

## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Making a lollipop chart for pinpercent. 

```{r}
ggplot(candy) +
aes(pricepercent, reorder(rownames(candy), pricepercent)) +
geom_segment(aes(yend = reorder(rownames(candy), pricepercent),
xend = 0), col="blue") +
geom_point()
```

# Exploring the correlation structure

```{r}
# install.packages('corrplot')
```

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
cij
```

```{r}
corrplot(cij, diag = F)
```

## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
chocolate and fruity (most negative), pluribus and bar. 

## Q23. Similarly, what two variables are most positively correlated?
chocolate and bar

# PCA

```{r}
pca <- prcomp(candy, scale. = T)
summary(pca)
```

```{r}
loadings <- as.data.frame(pca$rotation)
ggplot(loadings) +
aes(PC1,
reorder(rownames(loadings), PC1),
fill=PC1) +
geom_col()
```

Let's make our PC1 vs PC2 plot: 
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) +
aes(x=PC1, y=PC2,
size=winpercent/100,
text=rownames(my_data),
label=rownames(my_data)) +
geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7) +
theme(legend.position = "none") +
labs(title="Halloween Candy PCA Space", 
subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown)",
caption="Data from 538")
```
```{r}
#install.packages('plotly')
```

```{r}
library(plotly)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. 